---
title: "ST558 Project 1"
format: html
editor: visual
author: "Ryan Mersereau"
---

# Intro

Our goal for this project is to write functions that will manipulate and process data sets from a census API. We'll create *generic* functions to automatically summarize and plot certain returned data.

## Data

We are going to query the Public Use Microdata Sample (PUMS) Census API. Information about this is available here: [Information about PUMS API](https://www.census.gov/data/developers/data-sets/census-microdata-api.html)

# Part 1: Data Processing:

## Parsing Variable Values

First, we'll start by downloading and parsing the JSON files with variable descriptions for the years
2021-2024.

```{r}
#| warning: FALSE
#| message: FALSE
library(httr)
library(jsonlite)
library(tibble)
```

We only want to parse certain variables, so we'll specify them below.

We want to write a function to `GET()` the URL, parse the JSON, and pull out a list of variables with their values. We'll obtain one list for each year 

```{r}
vars <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP",
               "FER", "HHL", "SCH", "SCHL", "SEX", "REGION", "DIVISION")
# ST: 2021-2022
# STATE: 2023-2024

#API Key - Required, should find some way to encrypt this in future
key <- "4270e8325c6e236168e18ec3682a2a6cc27469c7"

# Function
get_var_dict <- function(year) {
  if (!(year  %in% 2021:2024)) {
    stop("Year must be between 2021-2024")
  }
  
  url <- paste0("https://api.census.gov/data/", year, "/acs/acs1/pums/variables.json?key=", key)
  response <- GET(url)
  stop_for_status(response)
  parsed <- fromJSON(content(response, as = "text", encoding = "UTF-8"),
                             simplifyVector = FALSE)
  
  state_var <- if (year <= 2022) "ST" else "STATE"
  
  #For each variable, keep a named character vector and sort by numeric order
  lapply(parsed$variables[c(vars, state_var)], function(vector) {
    items <- unlist(vector$values$item)
    code_num <- suppressWarnings(as.numeric(names(items)))
    items[order(is.na(code_num), code_num)]
  })
}

var_dicts <- lapply(2021:2024, get_var_dict)
names(var_dicts) <- 2021:2024
```

We can then save these lists as an `.rds` file in the repo under a data folder.
Using `dir.create()` creates the folder from scratch, and rerunning it without warnings
does nothing, allowing for more replicability.

```{r}
dir.create("data", showWarnings = FALSE)
saveRDS(var_dicts, "data/var_dicts.rds")
```

## Parsing API Data

```{r}
# Testing via a hard-coded query
response <- GET(
  "https://api.census.gov/data/2024/acs/acs1/pums",
  query = list(
    get   = "SEX,PWGTP,AGEP",
    `for` = "state:37",
    key   = key
  )
)

parsed <- fromJSON(content(response, as = "text", encoding = "UTF-8"))
head(parsed)
```

