---
title: "ST558 Project 1"
format: html
editor: visual
author: "Ryan Mersereau"
---

# Intro

Our goal for this project is to write functions that will manipulate and process data sets from a census API. We'll create *generic* functions to automatically summarize and plot certain returned data.

## Data

We are going to query the Public Use Microdata Sample (PUMS) Census API. Information about this is available here: [Information about PUMS API](https://www.census.gov/data/developers/data-sets/census-microdata-api.html)

# Part 1: Data Processing:

## Parsing Variable Values

First, we'll start by downloading and parsing the JSON files with variable descriptions for the years 2021-2024.

```{r}
#| warning: FALSE
#| message: FALSE
library(httr)
library(jsonlite)
library(tidyverse)
```

We only want to parse certain variables, so we'll specify them below.

We want to write a function to `GET()` the URL, parse the JSON, and pull out a list of variables with their values. We'll obtain one list for each year

```{r}
vars <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP",
               "FER", "HHL", "SCH", "SCHL", "SEX", "REGION", "DIVISION")
# ST: 2021-2022
# STATE: 2023-2024

#API Key - Required, should find some way to encrypt this in future
key <- "4270e8325c6e236168e18ec3682a2a6cc27469c7"

# Function
get_var_dict <- function(year) {
  if (length(year) != 1 || !(year %in% 2021:2024)) {
    stop("Year must be a single value between 2021 and 2024")
  }
  
  url <- paste0("https://api.census.gov/data/", year, "/acs/acs1/pums/variables.json?key=", key)
  response <- GET(url)
  stop_for_status(response)
  parsed <- fromJSON(content(response, as = "text", encoding = "UTF-8"), # Can also use the `rawToChar()` approach here 
                             simplifyVector = FALSE)
  
  state_var <- if (year <= 2022) "ST" else "STATE"
  
  #For each variable, keep a named character vector and sort by numeric order
  lapply(parsed$variables[c(vars, state_var)], function(vector) {
    items <- unlist(vector$values$item)
    code_num <- suppressWarnings(as.numeric(names(items)))
    items[order(is.na(code_num), code_num)]
  })
}

var_dicts <- lapply(2021:2024, get_var_dict)
names(var_dicts) <- 2021:2024
```

We can then save these lists as an `.rds` file in the repo under a data folder. Using `dir.create()` creates the folder from scratch, and rerunning it without warnings does nothing, allowing for more replicability.

```{r}
dir.create("data", showWarnings = FALSE)
saveRDS(var_dicts, "data/var_dicts.rds")
```

## Parsing API Data

Now, we'll want to write a function to query the PUMS API and return a nicely formatted tibble. When we query the API with our usual process, the data should come out nicely. Let's start with an example of this.

```{r}
# Testing via a hard-coded query
url <- paste0("https://api.census.gov/data/2024/acs/acs1/pums",
              "?get=SEX,PWGTP,AGEP&for=state:37",
              "&key=", key)
response <- GET(url)
parsed <- fromJSON(content(response, as = "text", encoding = "UTF-8"))
head(parsed)
```

Next, we'll write a *helper* function to take what is returned by `GET()` and turn it into a nice `tibble`

```{r}
response_to_tibble <- function(response) {
  stop_for_status(response)
  parsed <- fromJSON(content(response, as = "text", encoding = "UTF-8"))
  colnames(parsed) <- parsed[1, ]
  as_tibble(parsed[-1, , drop = FALSE]) # drop = FALSE keeps matrix shape for the tibble
}
# Testing on response from above
pums_tbl <- response_to_tibble(response)
str(pums_tbl)
```

Now, we'll wrap this into a function that also parses the data by writing a function to query the API that allows the user to change the following items:

- Year of survey (2024 as default)
  - Check a valid value given
- Specify numeric variables to be returned (AGEP and PWGTP as default)
  - Turn variables into numeric values or time (use `lubridate`)
- Specify the categorical variables to be returned (SEX as default)
  - Must include at least one categorical variable
  - Turn variables into factors with appropriate levels
- Specify geography level: Region, Division, or State
- Specify the specific region, division, or state to subset data on
  - Specify a default level if the user doesn't
  - Allowable values are in the `.rds` file
  - Allow input all option that would return all values for that geography
- Return an error if user specifies a variable not included in the above two lists

```{r}
query_pums <- function(year = 2024,
                       num_vars = c("AGEP", "PWGTP"),
                       cat_vars = "SEX",
                       geo_level = "State",
                       geo_value = NULL,
                       api_key = key) {
  # Stop if invalid year
  if (length(year) != 1 || !(year %in% 2021:2024)) {
    stop("year must be a single value: 2021, 2022, 2023, or 2024")
  }
  # Numeric variable options
  num_options <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
  num_vars <- toupper(num_vars)
  bad_num <- setdiff(num_vars, c(num_options, "PWGTP"))
  if (length(bad_num) > 0) {
    stop("Invalid numeric variable(s): ", paste(bad_num, collapse = ", "), # Stop if invalid numeric variables requested
         ". Choose from: ", paste(num_options, collapse = ", "))
  }
  if (!any(num_vars %in% num_options)) {
    stop("Must request at least one numeric variable (other than PWGTP)") #Stop if no numeric variables requested
  }
  num_vars <- union(num_vars, "PWGTP")  # PWGTP is always returned
  # Categorical variable options
  cat_options <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  cat_vars <- toupper(cat_vars)
  if (length(cat_vars) == 0) {
    stop("At least one categorical variable must be requested")
  }
  bad_cat <- setdiff(cat_vars, cat_options)
  if (length(bad_cat) > 0) {
    stop("Invalid categorical variable(s): ", paste(bad_cat, collapse = ", "), # Stop if invalid categorical variables requested
         ". Choose from: ", paste(cat_options, collapse = ", "))
  }
  # Geography level specification
  if (length(geo_level) != 1 ||
      !(tolower(geo_level) %in% c("region", "division", "state"))) {
    stop("geo_level must be one of 'Region', 'Division', or 'State'") # Stop if no Region, Division, or State given
  }
  geo_level <- tolower(geo_level)

  # Geography value options
  var_dicts <- readRDS("data/var_dicts.rds")
  geo_var <- switch(geo_level,
                    region   = "REGION",
                    division = "DIVISION",
                    state    = if (year <= 2022) "ST" else "STATE")
  geo_labels  <- var_dicts[[as.character(year)]][[geo_var]]
  valid_codes <- names(geo_labels)[!grepl("Puerto Rico", geo_labels)]

  # Defaults if no value given (chosen): South region (3), South Atlantic division (5), North Carolina (37)
  if (is.null(geo_value)) {
    geo_value <- switch(geo_level, region = "3", division = "5", state = "37")
  }

  if (length(geo_value) == 1 && tolower(geo_value) == "all") {
    geo_codes <- "*"
  } else {
    # Match integers so 1, "1", and "01" all work
    match_idx <- match(suppressWarnings(as.integer(geo_value)),
                       as.integer(valid_codes))
    if (anyNA(match_idx)) {
      stop("Invalid ", geo_level, " value(s): ", # Stop if invalid geo level
           paste(geo_value[is.na(match_idx)], collapse = ", "))
    }
    geo_codes <- paste(valid_codes[match_idx], collapse = ",")
  }
  # Query API and return a census tibble
  response <- GET(
    paste0("https://api.census.gov/data/", year, "/acs/acs1/pums"),
    query = list(get   = paste(c(num_vars, cat_vars), collapse = ","),
                 `for` = paste0(geo_level, ":", geo_codes),
                 key   = api_key)
  )
  pums_tbl <- response_to_tibble(response)

  class(pums_tbl) <- c("census", class(pums_tbl))
  pums_tbl
}  
```

Now lets test some queries that should work:

```{r}
query_pums()                                         # defaults: 2024, NC
query_pums(2022, num_vars = c("AGEP", "GASP"), cat_vars = c("SEX", "SCHL"))
# query_pums(2023, num_vars = "JWMNP", geo_level = "Region", geo_value = 1) # Takes a while
# query_pums(2024, geo_level = "Division", geo_value = "all") # Takes a while
```

These queries shouldn't work:

```{r}
#| error: true
query_pums(2020)                            # bad year
query_pums(num_vars = "PWGTP")              # no numeric var besides PWGTP
query_pums(cat_vars = "MAR")                # not an allowed categorical
query_pums(geo_level = "All")               # not allowed
query_pums(geo_level = c("State", "Region"))# multiple levels
query_pums(geo_value = 99)                  # not a valid state code
```

## Combining Years