---
title: "ST558 Project 1"
format: html
editor: visual
author: "Ryan Mersereau"
---

# Intro

Our goal for this project is to write functions that will manipulate and process data sets from a census API. We'll create *generic* functions to automatically summarize and plot certain returned data.

## Data

We are going to query the Public Use Microdata Sample (PUMS) Census API. Information about this is available here: [Information about PUMS API](https://www.census.gov/data/developers/data-sets/census-microdata-api.html)

# Part 1: Data Processing:

## Parsing Variable Values

First, we'll start by downloading and parsing the JSON files with variable descriptions for the years 2021-2024.

```{r}
#| warning: FALSE
#| message: FALSE
library(httr)
library(jsonlite)
library(tidyverse)
library(hms)
```

We only want to parse certain variables, so we'll specify them below.

We want to write a function to `GET()` the URL, parse the JSON, and pull out a list of variables with their values. We'll obtain one list for each year

```{r}
vars <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP",
               "FER", "HHL", "SCH", "SCHL", "SEX", "REGION", "DIVISION")
# ST: 2021-2022
# STATE: 2023-2024

#API Key - Required, should find some way to encrypt this in future
key <- "4270e8325c6e236168e18ec3682a2a6cc27469c7"

# Function
get_var_dict <- function(year) {
  if (length(year) != 1 || !(year %in% 2021:2024)) {
    stop("Year must be a single value between 2021 and 2024")
  }
  
  url <- paste0("https://api.census.gov/data/", year, "/acs/acs1/pums/variables.json?key=", key)
  response <- GET(url)
  stop_for_status(response)
  parsed <- fromJSON(content(response, as = "text", encoding = "UTF-8"), # Can also use the `rawToChar()` approach here 
                             simplifyVector = FALSE)
  
  state_var <- if (year <= 2022) "ST" else "STATE"
  
  #For each variable, keep a named character vector and sort by numeric order. Non-numeric codes become NA
  lapply(parsed$variables[c(vars, state_var)], function(variable) {
    items <- unlist(variable$values$item)
    code_num <- suppressWarnings(as.numeric(names(items)))
    items[order(code_num)]
  })
}

var_dicts <- lapply(2021:2024, get_var_dict)
names(var_dicts) <- 2021:2024
```

We can then save these lists as an `.rds` file in the repo under a data folder. Using `dir.create()` creates the folder from scratch, and rerunning it without warnings does nothing, allowing for more replicability.

```{r}
dir.create("data", showWarnings = FALSE)
saveRDS(var_dicts, "data/var_dicts.rds")
```

## Parsing API Data

Now, we'll want to write a function to query the PUMS API and return a nicely formatted tibble. When we query the API with our usual process, the data should come out nicely. Let's start with an example of this.

```{r}
# Testing via a hard-coded query, using state NC
url <- paste0("https://api.census.gov/data/2024/acs/acs1/pums",
              "?get=SEX,PWGTP,AGEP&for=state:37",
              "&key=", key)
response <- GET(url)
parsed <- fromJSON(content(response, as = "text", encoding = "UTF-8"))
head(parsed)
```

Next, we'll write a *helper* function to take what is returned by `GET()` and turn it into a nice `tibble`

```{r}
response_to_tibble <- function(response) {
  stop_for_status(response)
  parsed <- fromJSON(content(response, as = "text", encoding = "UTF-8"))
  colnames(parsed) <- parsed[1, ]
  as_tibble(parsed[-1, , drop = FALSE]) # drop = FALSE keeps matrix shape for the tibble
}
# Testing on response from above
pums_tbl <- response_to_tibble(response)
str(pums_tbl)
```

Then, let's make some more helper functions to help us with our next step. These will convert variables into numeric values, normalize time codes and return the midpoint, and handle categorical variables by making them factors.

```{r}
# Strip leading zeros to help standard time codes: "01", "1", and "001" all match
normalize_code <- function(x) {
   sub("^0+(?=.)", "", x, perl = TRUE) }

# Convert to numbers; codes labeled "N/A" in the dictionary become NA
convert_numeric <- function(x, labels) {
  na_codes <- normalize_code(names(labels)[grepl("^N/A", labels)])
  x_num <- suppressWarnings(as.numeric(x))
  x_num[normalize_code(x) %in% na_codes] <- NA
  x_num
}

# Time: midpoint of each interval as a time of day ("N/A" labels become NA)
convert_time <- function(x, labels) {
  clean <- gsub("\\.", "", labels)   # "a.m." -> "am"
  start <- parse_date_time(sub(" to .*", "", clean), "I:M p", quiet = TRUE)
  end   <- parse_date_time(sub(".* to ", "", clean), "I:M p", quiet = TRUE)
  midpoint <- as_hms(start + (end - start) / 2)
  midpoint[match(normalize_code(x), normalize_code(names(labels)))]
}

# Categorical: factor with dictionary-ordered levels and readable labels
make_factor <- function(x, labels) {
  factor(normalize_code(x), levels = normalize_code(names(labels)), labels = labels)
}
```

Now, using our helper functions, we'll wrap this into a function that also parses the data by writing a function to query the API that allows the user to change the following items:

- Year of survey (2024 as default)
  - Check a valid value given
- Specify numeric variables to be returned (AGEP and PWGTP as default)
  - Turn variables into numeric values or time (use `lubridate`)
- Specify the categorical variables to be returned (SEX as default)
  - Must include at least one categorical variable
  - Turn variables into factors with appropriate levels
- Specify geography level: Region, Division, or State
- Specify the specific region, division, or state to subset data on
  - Specify a default level if the user doesn't
  - Allowable values are in the `.rds` file
  - Allow input all option that would return all values for that geography
- Return an error if user specifies a variable not included in the above two lists

```{r}
query_pums <- function(year = 2024,
                       num_vars = c("AGEP", "PWGTP"),
                       cat_vars = "SEX",
                       geo_level = "State",
                       geo_value = NULL,
                       api_key = key) {
  # Stop if invalid year
  if (length(year) != 1 || !(year %in% 2021:2024)) {
    stop("year must be a single value: 2021, 2022, 2023, or 2024")
  }
  # Numeric variable options
  num_options <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
  num_vars <- toupper(num_vars)
  bad_num <- setdiff(num_vars, c(num_options, "PWGTP"))
  if (length(bad_num) > 0) {
    stop("Invalid numeric variable(s): ", paste(bad_num, collapse = ", "), # Stop if invalid numeric variables requested
         ". Choose from: ", paste(num_options, collapse = ", "))
  }
  if (!any(num_vars %in% num_options)) {
    stop("Must request at least one numeric variable (other than PWGTP)") 
  }
  num_vars <- union(num_vars, "PWGTP")  # PWGTP is always returned
  # Categorical variable options
  cat_options <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  cat_vars <- toupper(cat_vars)
  if (length(cat_vars) == 0) {
    stop("At least one categorical variable must be requested")
  }
  bad_cat <- setdiff(cat_vars, cat_options)
  if (length(bad_cat) > 0) {
    stop("Invalid categorical variable(s): ", paste(bad_cat, collapse = ", "), 
         ". Choose from: ", paste(cat_options, collapse = ", "))
  }
  # Geography level specification
  if (length(geo_level) != 1 ||
      !(tolower(geo_level) %in% c("region", "division", "state"))) {
    stop("geo_level must be one of 'Region', 'Division', or 'State'") 
  }
  geo_level <- tolower(geo_level)

  # Geography value options
  var_dicts <- readRDS("data/var_dicts.rds")
  geo_var <- switch(geo_level,
                    region   = "REGION",
                    division = "DIVISION",
                    state    = if (year <= 2022) "ST" else "STATE")
  geo_labels  <- var_dicts[[as.character(year)]][[geo_var]]
  valid_codes <- names(geo_labels)[!grepl("Puerto Rico", geo_labels)]

  # Defaults if no value given (chosen): South region (3), South Atlantic division (5), North Carolina (37)
  if (is.null(geo_value)) {
    geo_value <- switch(geo_level, region = "3", division = "5", state = "37")
  }

  if (length(geo_value) == 1 && tolower(geo_value) == "all") {
    geo_codes <- "*"
  } else {
    # Match integers so 1, "1", and "01" all work
    match_idx <- match(suppressWarnings(as.integer(geo_value)),
                       as.integer(valid_codes))
    if (anyNA(match_idx)) {
      stop("Invalid ", geo_level, " value(s): ", 
           paste(geo_value[is.na(match_idx)], collapse = ", "))
    }
    geo_codes <- paste(valid_codes[match_idx], collapse = ",")
  }
  # Query API and return a census tibble
  response <- GET(
    paste0("https://api.census.gov/data/", year, "/acs/acs1/pums"),
    query = list(get   = paste(c(num_vars, cat_vars), collapse = ","),
                 `for` = paste0(geo_level, ":", geo_codes),
                 key   = api_key)
  )
  pums_tbl <- response_to_tibble(response)
  
  # Convert column types using this years dictionary
  dict <- var_dicts[[as.character(year)]]
  for (v in num_vars) {
    pums_tbl[[v]] <- if (v %in% c("JWAP", "JWDP")) {
      convert_time(pums_tbl[[v]], dict[[v]])
    } else {
      convert_numeric(pums_tbl[[v]], dict[[v]])
    }
  }
  for (v in cat_vars) {
    pums_tbl[[v]] <- make_factor(pums_tbl[[v]], dict[[v]])
  }
  # droplevels() removes geographies that weren't requested
  pums_tbl[[geo_level]] <- droplevels(make_factor(pums_tbl[[geo_level]], dict[[geo_var]]))

  class(pums_tbl) <- c("census", class(pums_tbl))
  pums_tbl
}  
```

Now lets test our function:

```{r}
# Default values, year = 2024, State = NC
query_pums()                                         
```

```{r}
# Specified values
nc_2022_test <- query_pums(2022,
                        num_vars = c("AGEP", "JWAP", "JWMNP"),
                        cat_vars = c("SEX", "SCHL"))
str(nc_2022_test)
```

These queries shouldn't work:

```{r}
#| error: TRUE

query_pums(2020)                            # bad year
query_pums(num_vars = "PWGTP")              # no numeric var besides PWGTP
query_pums(cat_vars = "MAR")                # not an allowed categorical
query_pums(geo_level = "All")               # not allowed
query_pums(geo_level = c("State", "Region"))# multiple levels
query_pums(geo_value = 99)                  # not a valid state code
```

## Combining Years

Finally, lets write a function that allows the user to specify multiple years of survey data (along with all the other options). This function will call a single year function as many times as needed and then combine the data into one final tibble with a year variable included.

```{r}
query_pums_combined <- function(years = 2024, ...) { # ... passes everything else to query_pums (not repeated)
  years <- sort(unique(years))
  if (length(years) == 0 || !all(years %in% 2021:2024)) {
    stop("Years must be one or more of: 2021, 2022, 2023, 2024")
  }

  # Query each year, add a year column, and drop the census class before binding
  results <- lapply(years, function(yr) {
    year_tbl <- query_pums(year = yr, ...)
    year_tbl$year <- yr
    class(year_tbl) <- setdiff(class(year_tbl), "census")
    year_tbl
  })

  combined <- bind_rows(results)
  combined$year <- factor(combined$year)

  class(combined) <- c("census", class(combined))
  combined
}
```

Testing a combined years query

```{r}
multi <- query_pums_combined(c(2022, 2024),
                    num_vars = c("AGEP", "JWMNP"),
                    cat_vars = "SEX")
str(multi)
class(multi)          
```

A year outside 2021-2024 should still return an error

```{r}
#| error: TRUE

query_pums_combined(c(2021, 2020))
```

# Part 2: Writing a Generic Function for Summarizing

The rows in our tibbles don't generally correspond to a single person! The `PWGTP` variable actually represents the number of (people) observations there should be for a particular row.

We could create a new tibble with each row replicated `PWGTP` times, but this would create a very large tibble in some cases.

Instead, we'll use **Generic Functions**

Since we have added a `census` class to our tibble, we can write our own custom `summary()` function!

## Numeric Summaries

For the `census` summary method, let's write a function that produces means and standard deviations for our numeric variable(s) and counts for our categorical variable(s)

- The function will take three arguments: the census tibble (maybe having multiple years), the numeric variable(s) to summarize, and the categorical variable(s) to summarize.

- By default, it will summarize all numeric variables (other than `PWGTP`) and all categorical variables in the tibble. However, the user will be able to specify the variables they'd like to summarize.

- All values will be returned as a named list

```{r}
summary.census <- function(object, num_vars = NULL, cat_vars = NULL, ...) {
  weight <- object$PWGTP

  # Defaults: all numeric variables (except PWGTP) and all factors
  if (is.null(num_vars)) {
    num_vars <- names(object)[sapply(object, is.numeric)]
    num_vars <- setdiff(num_vars, "PWGTP")
  }
  if (is.null(cat_vars)) {
    cat_vars <- names(object)[sapply(object, is.factor)]
  }

  # Validate user input
  missing_vars <- setdiff(c(num_vars, cat_vars), names(object))
  if (length(missing_vars) > 0) {
    stop("Variable(s) not in the data: ", paste(missing_vars, collapse = ", "))
  }
  if (!all(sapply(object[num_vars], is.numeric))) {
    stop("num_vars must be numeric, non-time variables")
  }
  if (!all(sapply(object[cat_vars], is.factor))) {
    stop("cat_vars must be categorical (factor) variables")
  }

  # Weighted mean and SD for each numeric variable, dropping NAs
  num_summaries <- lapply(num_vars, function(v) {
    keep <- !is.na(object[[v]])
    x <- object[[v]][keep]
    w <- weight[keep]
    mean_x <- sum(x * w) / sum(w)
    sd_x <- sqrt(sum(x^2 * w) / sum(w) - mean_x^2)
    c(mean = mean_x, sd = sd_x)
  })
  names(num_summaries) <- num_vars

  # Weighted counts for each categorical variable
  cat_summaries <- lapply(cat_vars, function(v) {
    tapply(weight, object[[v]], sum, default = 0)
  })
  names(cat_summaries) <- cat_vars

  list(numeric = num_summaries, categorical = cat_summaries)
}
```

Now, lets test our function in several ways

```{r}
# Default values (all numeric except PWGTP)
summary(multi)
```

```{r}
# One numeric, two categorical
summary(multi, num_vars = "JWMNP", cat_vars = c("year", "SEX"))
```

Bad example: Trying to pass time variables

```{r}
#| error: TRUE

summary(multi, num_vars = "JWAP")
```

## Graphical Summaries

Similar to this process, let's create a generic `plot()` function for a `census` class tibble. This will require the user to specify one categorical variable (can include year) and one numeric variable for plotting. The plot should create a boxplot of the numeric variable at the levels of the categorical variable.

Quickly, let's also define some of the labels for our variables so our plots are more readable

```{r}
# Readable labels for plot titles and axes
var_labels <- c(
  AGEP     = "Age (years)",
  GASP     = "Monthly gas cost ($)",
  GRPIP    = "Gross rent as % of household income",
  JWAP     = "Time of arrival at work",
  JWDP     = "Time of departure for work",
  JWMNP    = "Travel time to work (minutes)",
  FER      = "Gave birth in past 12 months",
  HHL      = "Household language",
  SCH      = "School enrollment",
  SCHL     = "Educational attainment",
  SEX      = "Sex",
  year     = "Survey year",
  state    = "State",
  region   = "Region",
  division = "Division"
)

# Return the readable label if one exists, otherwise the variable name
get_label <- function(var) {
  if (var %in% names(var_labels)) var_labels[[var]] else var
}
```

Now, let's build our generic plot function

```{r}
plot.census <- function(x, cat_var, num_var, colors = NULL, ...) {
  # Both variables are required
  if (missing(cat_var) || missing(num_var)) {
    stop("You must specify one categorical (cat_var) and one numeric (num_var) variable")
  }
  if (length(cat_var) != 1 || length(num_var) != 1) {
    stop("cat_var and num_var must each be a single variable name")
  }

  # Validate the categorical variable
  if (!(cat_var %in% names(x)) || !is.factor(x[[cat_var]])) {
    stop(cat_var, " is not a categorical (factor) variable in the data")
  }

  # Validate the numeric variable (time values are allowed)
  if (!(num_var %in% names(x)) ||
      !(is.numeric(x[[num_var]]) || inherits(x[[num_var]], "hms"))) {
    stop(num_var, " is not a numeric variable in the data")
  }
  if (num_var == "PWGTP") {
    stop("PWGTP is the weight variable; choose a different numeric variable")
  }

  # Drop rows with missing values, drop unused factor levels
  plot_data <- x[!is.na(x[[num_var]]) & !is.na(x[[cat_var]]), ]
  plot_data[[cat_var]] <- droplevels(plot_data[[cat_var]])

  p <- ggplot(plot_data, aes(x = get(cat_var), y = get(num_var),
                             weight = PWGTP, fill = get(cat_var))) +
    geom_boxplot(...) +                                    # pass extra options through
    labs(x = get_label(cat_var),                           # readable labels
         y = get_label(num_var),
         title = paste(get_label(num_var), "by", get_label(cat_var))) +
    theme(legend.position = "none")                        # x-axis already labels the boxes

  # Apply user-chosen colors, one per level of the categorical variable
  if (!is.null(colors)) {
    n_levels <- nlevels(plot_data[[cat_var]])
    if (length(colors) < n_levels) {
      stop("Need at least ", n_levels, " colors for the levels of ", cat_var)
    }
    p <- p + scale_fill_manual(values = colors)
  }

  p
}
```

Now, let's test our plot function in several ways

```{r}
# One categorical variable: year and numeric variable age
plot(multi, cat_var = "year", num_var = "AGEP", colors = c("gold", "forestgreen"))
```

Here is a boxplot of age in years by survey years. We see very little change here which is expected as the survey years are only two years apart.

```{r}
# Categorical variable sex and numeric variable commute time
plot(multi, cat_var = "SEX", num_var = "JWMNP", colors = c("skyblue", "magenta"))
```

Here is a plot of travel time to work (in minutes) by sex. Both boxplots are centered around 25 minutes, with women having a slightly larger IQR for commute times. There are also some clear outliers for commuters over and hour and a few 'supercommuters' over two hours

```{r}
# Plot of time of arrival and sex with outliers lightened
plot(nc_2022_test, cat_var = "SEX", num_var = "JWAP", outlier.alpha = 0.05) # default colors
```
It's important to note that we don't use multi here because we are dealing with time, so we used our test query from earlier for 2022 NC data.

These boxplots of time of arrival by sex have much more variance due to many workers who work night shifts or atypical hours, but the median of the data is still around 9AM, with females averaging slightly later arrival, from this data.

Now, let's try a multi-level example of age by educational attainment using the same 2022 NC Data. Here we flip the coordinates for better readability since there are so many levels of education.

```{r}
plot(nc_2022_test, cat_var = "SCHL", num_var = "AGEP") +
  coord_flip()
```

Here we see a clear trend of education level and age being positively correlated. There are a lot of outliers for older people who never received a formal education beyond a few grade levels.

Finally, here are some error cases that shouldn't work

```{r}
#| error: true

plot(multi, cat_var = "year")                                         # missing num_var
plot(multi, cat_var = "AGEP", num_var = "SEX")                        # variables swapped
plot(multi, cat_var = "SCHL", num_var = "AGEP", colors = c("blue", "red"))  # too few colors
```

## Investigation: Changes in commuting trends

Continuing with some of the ideas above, let's investigate how commuting times changed from 2021 to 2024 in the state with one of the longest average commute times, New York. 2021 was still at the peak of COVID, so we will see if commute times shifted after this as more and more people got back on the roads and in-person work environments.

Lets start with a new pull of the data using our combined query function, selecting relevant numeric and categorical variables:

(We can also include the cache: TRUE option to save the query results on renders, which saves time since this is a large amount of data)

```{r}
#| cache: TRUE
commute_ny <- query_pums_combined(2021:2024,
                                  num_vars = c("JWMNP", "JWAP", "AGEP"),
                                  cat_vars = c("SEX", "SCHL"), geo_value = 36)
```

Next, let's display summaries by year for only commute times

```{r}
summary(commute_ny, num_vars = "JWMNP", cat_vars = "year")

```

Our numeric summary function only gives us pooled statistics. The mean one-way commute time across all four years in New York was 32.6 minutes. The categorical variable is total number of observations. This is useful, but we'll have to tweak this to get per year commute summaries

```{r}
commute_by_year <- sapply(levels(commute_ny$year), function(yr) {
  year_data <- commute_ny[commute_ny$year == yr, ]
  summary(year_data, num_vars = "JWMNP", cat_vars = "year")$numeric$JWMNP
})
commute_by_year
```

Here we can see mean and standard deviation commute times for the state by year. In 2021, the mean was the lowest, at 31.39 minutes. This mean increased to 33 minutes in 2022 and stayed around that length for the next two years.

Now, let's plot this commute length and arrival time by year

```{r}
plot(commute_ny, cat_var = "year", num_var = "JWMNP", outlier.alpha = 0.1)
plot(commute_ny, cat_var = "year", num_var = "JWAP", outlier.alpha = 0.05)
```

This boxplot supports our numerical summary findings of the average commute time increasing after the year 2021, although the IQR doesn't appear to move much. It's important to note that the center of the boxplot is the median, not mean, so median commute times gradually rose in 2022, and then substantially rose in 2023. The mean commute time for this data is higher than the median because the commute times are right-skewed.

Also, the time of arrival at work appears to have little change over the years.

Finally, lets compare New York to some other states: North Carolina and Texas in the year 2024 to see if there is a significant difference in commute times.

```{r}
commute_states <- query_pums(2024,
                             num_vars = c("JWMNP", "AGEP"),
                             cat_vars = "SEX",
                             geo_value = c(36, 37, 48))   # NY, NC, TX

summary(commute_states, num_vars = "JWMNP", cat_vars = "state")
plot(commute_states, cat_var = "state", num_var = "JWMNP",
     colors = c("steelblue", "darkorange", "darkred"))
```

Here we can see New York having a much higher mean travel time and Q3 than NC and Texas do for the year 2024.
